Union and intersection contracts are hard, actually

### 2021-06-09 paper.tex
Acquire lock on sec 6
Rework of section 6 (related works)

Like in the case of lazy dictionaries, the checks are delayed.
Contracts keeps track of whether the caller or the implementation is
at fault for a violation, hence it can report precise error
messages. Compare Figure~\ref{fig:contract-reporting-wo} and
Figure~\ref{fig:contract-reporting-w}: in
Figure~\ref{fig:contract-reporting-wo} an error is reported inside the
\nickel{catHosts} function, but \nickel{catHosts} is, in fact,
correct, as is made clear by Figure~\ref{fig:contract-reporting-w},
where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
contract, and correctly reports that the caller failed to call
\nickel{catHosts} with a string argument.

@@

Keil and Thiemann realizes the tour de force of combining union and intersection
contracts with user-defined contracts. The price to pay, though, is that
implementing their solution in a reasonable way is far from obvious. Not only
when starting from the more abstract, non-deterministic formulation, but even
when starting from the algorithmic version of their system.

@@

\subsection{Gradual typing: a new perspective}

Castagna et. al \cite{gradualCastagna} present a gradual type system that
incorporates multiple non-trivial typing constructs, in particular, subtyping,
unions, recursive types, negation types, and a bottom type.
Which can easily encode intersection, difference, and top types.

This system, even if related to the ones we studied, adheres to a different
school of gradual typing, Greenberg would call it the static first
school\cite{practiceTheoryGrad}.
In this line of work, the goal is usually to develop a type system that can
handle a dynamic type to represent the lack of precission on some type.
The main reasons why we consider this work not comparable with what we attempted
to solve are:
\begin{itemize}
  \item Like Williams et al\cite{RootCauseOfBlame}, it doesn't allow
        to use user defined contracts.
  \item Their base calculus needs to have a non-trivial type inference
        algorithm, that makes the development of a programming language
        much more cumbersome. For instance, it defines a set-theoretic
        interpretation of types, and a subtyping relation based on it.
  \item Contracts are completely handled by the execution engine (in the paper,
        the rewriting system), and not added as a library with some basic
        syntactic support.
\end{itemize}

@@

When trying to enforce properties on dynamic languages, unions appear naturally,
since they allow to express that a variable is in an unknown state, but within
some bounded set of alternatives.
Sadly, the implementation of union and intersection contracts is not as simple as
it first appears. In conjunction with higher-order contracts, or even just basic
contracts in a lazy language, they behave in a way that is reminiscent of
side-effects. These behaviors were, for most of them, already known: we do not
claim their discovery. However, as language implementers, we have made an actual
attempt at including them in a concrete language. We exposed what, in our
opinion, are the most important barriers and their impact on the rest of the
language. Even in a pure setting, union and intersection contracts break the
equivalence between call-by-name and call-by-value, and can consequently prevent
optimizations. There are discrepancies between the expected and actual semantics
that make some behaviors unintuitive. Their implementation is a non-trivial
change that contaminates the evaluation of all other contracts, and the
performance penalty is not obvious to assess.

>>>>

Like in the case of lazy dictionaries, the checks are delayed.
Contracts keep track of whether the caller or the implementation is
at fault for a violation, hence it can report precise error
messages. Compare Figure~\ref{fig:contract-reporting-wo} and
Figure~\ref{fig:contract-reporting-w}: in
Figure~\ref{fig:contract-reporting-wo} an error is reported inside the
\nickel{catHosts} function, but \nickel{catHosts} is, in fact,
correct, as is made clear by Figure~\ref{fig:contract-reporting-w},
where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
contract, and correctly reports that the caller failed to call
\nickel{catHosts} with a string argument.

@@

Keil and Thiemann realize the tour de force of combining union and intersection
contracts with user-defined contracts. The price to pay, though, is that
implementing their solution in a reasonable way is far from obvious. Not only
when starting from the more abstract, non-deterministic formulation, but even
when starting from the algorithmic version of their system.

@@

\subsection{Higher-order contracts}

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce higher-order contracts, a principled approach to run-time
assertion checking that nicely supports functions. They introduce the notion of
blame, which is crucial to good error reporting. It became apparent later that
their contracts are closely related to the type casts introduced by gradual
typing, excluding blame: both \cite{FindlerMultiLang} and
\cite{FelleisenInterLang} see the value of contracts as a safe interface between
typed and untyped code. In \cite{WellTypedBlamed}, the authors precisely
introduce a system integrating gradual typing with contracts \textit{à la
Findler \& Felleisen}. Nickel adopts a similar type system, with both statically
typed terms, dynamically typed terms, and first-class contracts. Higher-order
contracts are the basis of the work~\cite{KeilThiemannUnionIntersection}
and~\ref{RootCauseOfBlame} that this paper studied in detail.

\subsection{Unions and intersections in gradual typing}

Castagna et. al \cite{gradualCastagna} introduce a gradual type system based on
so-called set-theoretic types~\cite{SetTheoreticTypes}.  Set-theoretic types
feature unions, intersections, negation types together with a notion of
subtyping. This work adheres to the \emph{static first school} (see
~\cite{practiceTheoryGrad}) of gradual typing: the reason for gradual types is
to allow for a less precise type information. It follows that their goals and
constraints are somehow different, resulting in the absence of first-class
contracts. As for any gradual type system, they do have to implement casts -
that are very close to contracts - for unions and intersections, but these casts
are neither visible nor available to the programmer. Obviously, no contracts
mean no user-defined contracts as well.

The implementation of casts relies on abstract interpretation, leading to a
complex specification, whose implementation is straightforward. It is not clear
either how much these casts rely on static type information to function
correctly, possibly hampering their transformation to more general standalone
contracts.

@@

When trying to enforce properties on dynamic languages, unions appear naturally,
since they allow to express that a variable is in an unknown state, but within
some bounded set of alternatives.  Sadly, the implementation of union and
intersection contracts is not as simple as it first appears. In conjunction with
higher-order contracts, they introduce side-effects in the language. These
behaviors were, for most of them, already known: we do not claim their
discovery. However, as language implementers, we have made an actual attempt at
including them in a concrete language. We exposed what, in our opinion, are the
most important barriers and their impact on the rest of the language. Even in a
pure setting, union and intersection contracts break the equivalence between
call-by-name and call-by-value, and can consequently prevent optimizations.
There are discrepancies between the expected and actual semantics that make some
behaviors unintuitive. Their implementation is a non-trivial change that
contaminates the evaluation of all other contracts, and the performance penalty
is not obvious to assess.
